Effect of initial volume fraction on the collapse of granular columns in fluid

### 2018-08-26 paper/paper.tex
:pencil: First draft

% packages needed
\usepackage{epsfig}
\usepackage{timesmt}

% add here more packages based on the document format
% custom packages
\usepackage{dsfont}
\usepackage{amsmath}
\setlength{\mathindent}{0cm}
\usepackage{subfig}
\usepackage{siunitx}
\usepackage[noabbrev]{cleveref}
\usepackage[font=small,justification=justified,singlelinecheck=false]{caption}

@@

\subsection{Effect of initial density}

>>>>

% packages needed
%\usepackage{subfigure}
\usepackage{epsfig}
\usepackage{timesmt}
\mathindent=0pt%

% custom packages
\usepackage{amsmath}
\usepackage[noabbrev]{cleveref}
\usepackage{dsfont}
\setlength{\mathindent}{0cm}
%\usepackage{subfig}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{siunitx}
\usepackage[font=small,justification=justified,singlelinecheck=false]{caption}

@@

\subsection{Effect of initial packing density}

\citeNP{Rondon2011} observed that the loose packings flow rapidly on a time
scale proportional to the initial height and results in longer run-out distance 
in comparison to the dense packing. Hydroplaning occurs above a critical 
Froude's number of 0.4. The Froude's number is inversely related to the 
thickness of the flow and its density. Hence, for the same thickness of flow, a 
loose granular column will experience more hydroplaning than a dense granular 
flow. This effect might result in longer run-out behaviour in fluid than the 
dry condition for the same initial aspect ratio. The initial packing density 
and the permeability of a 2D granular column, with an aspect ratio of 0.8, are 
varied to understand their influence on the run-out behaviour. The run-out 
behaviour of the dense case (83\% packing density), discussed in the previous 
section, is compared with a loose granular column (79\% packing fraction). The 
permeability is varied by changing the hydrodynamic radius from 0.7 \textit{R} 
to 0.95 
R.

The normalised run-out evolution with time for a loose initial packing (79\% 
packing fraction) with different hydrodynamic radii 0.7 \textit{R}, 0.8 
\textit{R}, 0.9 and 0.95 R are presented in~\cref{fig:Runout_a08_loose}. The 
run-out evolution a column of grains in suspension is compared with the dry and 
submerged granular columns to understand the influence of hydrodynamic forces 
on the flow kinematics. Similar to the dense granular column, the run-out 
distance increases with increase in the hydrodynamic radius (i.e., decrease in 
permeability). At low permeabilities (\textit{r} = 0.9 and 0.95 \textit{R}), 
the run-out distance is longer than the dry condition. This shows that the 
lubrication effect in low permeability conditions overcomes the influence of 
the drag force and the development of large negative pore-pressure resulting in 
a longer run-out distance. Although the suspended granular masses experience 
higher drag forces and turbulent effects, the run-out evolves almost at the 
same rate in comparison with granular columns with high permeability. This 
shows the effect of permeability on the dissipation rate of negative 
pore-pressure developed during the initial stage of collapse.

\begin{figure}
\centering
\includegraphics[width=0.9\linewidth]{figs/Runout_a08_loose.pdf}
\caption{Effect of permeability on the evolution of run-out for a column 
collapse in fluid (a = 0.8 \& loose packing).}
\label{fig:Runout_a08_loose}
\end{figure}

\Cref{fig:Loose_PWP_ini} shows the development of negative pore-pressure in low 
permeability (\textit{r} = 0.95 \textit{R}) and dissipation of negative 
pore-pressure in high permeability (\textit{r} = 0.7 \textit{R}) at the same 
time $ t = \tau_c$. This difference in the quantity and the rate of dissipation 
of negative pore-pressure results in a difference in the rate of flow 
evolution. A low permeability column requires a longer duration to 
evolve.~\Cref{fig:Loose_PWP_flow} shows the distribution of the excess 
pore-pressure along the bottom for low and high permeability conditions. As the 
flow progresses, the low permeability of the granular column causes 
hydroplaning to occur at the base of the column, which can be observed by high 
positive pore-pressure at the base of the flow front 
(\cref{fig:Loose_r095_PWP_flow,fig:Loose_r095_PWPress_flow}), resulting in a 
longer run-out distance.

\begin{figure}
\centering
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/Loose_r07_PWP_ini}
    \caption{High permeability (\textit{r} = 0.7 \textit{R}) - Pressure at the 
        bottom of the granular flow.}
    \label{fig:Loose_r07_PWP_ini}
\end{subfigure}
}\\
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/Loose_r095_PWP_ini}
    \caption{Low permeability (\textit{r} = 0.95 \textit{R}) - Pressure at the 
    bottom of the granular flow.}
    \label{fig:Loose_r095_PWP_ini}
\end{subfigure}
}
\end{figure}
\begin{figure}
\ContinuedFloat
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r07_PWP_ini_press_loose}
    \caption{High permeability (\textit{r} = 0.7 \textit{R}) - Pressure contour 
    (Pa).}
    \label{fig:Loose_r07_PWPress_ini}
\end{subfigure}
}\\
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r095_PWP_ini_press_loose}
    \caption{Low permeability (\textit{r} = 0.95 \textit{R}) - Pressure contour 
    (Pa).}
    \label{fig:Loose_r095_PWPress_ini}
\end{subfigure}
}
\caption{Effect of permeability on the excess pore water pressure distribution 
along the base of a granular column collapse in fluid (a = 0.8 \& loose 
packing) at $t = \tau_c$.}
\label{fig:Loose_PWP_ini}
\end{figure}

\begin{figure}
\centering
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/Loose_r07_PWP_flow}
    \caption{High permeability (\textit{r} = 0.7 \textit{R}) - Pressure at the 
        bottom of the granular flow.}
    \label{fig:Loose_r07_PWP_flow}
\end{subfigure}
}\\
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/Loose_r095_PWP_flow}
    \caption{Low permeability (\textit{r} = 0.95 \textit{R}) - Pressure at the 
        bottom of the granular flow.}
    \label{fig:Loose_r095_PWP_flow}
\end{subfigure}
}
\end{figure}
\begin{figure}
\ContinuedFloat
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r07_PWP_flow_press_loose}
    \caption{High permeability (\textit{r} = 0.7 \textit{R}) - Pressure contour 
    (Pa).}
    \label{fig:Loose_r07_PWPress_flow}
\end{subfigure}
}\\
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r095_PWP_flow_press_loose}
    \caption{Low permeability (\textit{r} = 0.95 \textit{R}) - Pressure contour 
    (Pa).}
    \label{fig:Loose_r095_PWPress_flow}
\end{subfigure}
}
\caption{Effect of permeability on the excess pore water pressure distribution 
for a granular column collapse in fluid (a = 0.8 \& loose packing) at $t = 
2\tau_c$.}
\label{fig:Loose_PWP_flow}
\end{figure}

The evolution of the potential energy with time (\cref{fig:PE_a08_loose}) 
reveals that at a very low permeability (\textit{r} = 0.95 \textit{R}), the 
initial potential energy mobilised is smaller than at \textit{r} = 0.9 
\textit{R}. Also with decreasing permeability, the time required to 
dissipate the negative pore-pressure increases. This results in a shorter 
run-out distance in the case of \textit{r} = 0.95 \textit{R} to that of 
\textit{r} = 0.9 \textit{R}. As the quantity of material destabilised is small, 
the flow is thinner and thus has a high Froude's number 
(0.59).~\Cref{fig:KExy_a08_loose} shows that the peak horizontal kinetic 
velocity observed in the case of \textit{r} = 0.9 \textit{R} is higher than 
\textit{r} = 0.95 \textit{R}. A Froude's number of 0.59 for \textit{r} = 0.9 
\textit{R} is observed in contrast to 0.46 for \textit{r} = 0.95 \textit{R}. 
Both values of hydrodynamic radii result in a Froude's number that indicate the 
occurrence of hydroplaning. However, the difference in the amount of material 
destabilised for \textit{r} = 0.95 \textit{R} and the decreased effect of 
hydroplaning results in a shorter run-out distance for \textit{r} = 0.95 
\textit{R} in comparison to \textit{r} = 0.9 \textit{R}.

\Cref{fig:Packing_Density_a08_loose} shows the evolution of packing 
fraction with time for different values of permeability. As the column 
collapses, water is entrained at the flow front. This can 
observed by the decrease in the packing fraction during $t = \tau_c$ and $t = 
3\tau_c$. As the flow progresses, the entrained water is expelled and the soil 
grains consolidate to reach a critical packing density at the end of the flow. 
The permeability (i.e., hydrodynamic radius) plays a crucial role in the rate 
of dissipation of the entrained water. As the permeability decreases, the water 
entrained at the flow front takes longer time to be dissipated resulting in 
lubrication of the flow at low permeabilities.~\Cref{fig:Fr_a08_loose} shows 
that the low permeable columns exhibit higher Froude's numbers, greater than 
0.4, that indicates occurrence of hydroplaning. This lubrication effect results 
in an increase in the run-out distance for columns with low permeabilities.

\begin{figure}
	\centering
    \includegraphics[width=0.8\linewidth]{figs/PE_a08_loose}
    \caption{Effect of permeability on the evolution of the potential energy 
    with time for a granular column collapse in fluid (a = 0.8 \& loose 
    packing).}
    \label{fig:PE_a08_loose}
\end{figure}

\begin{figure}
\centering
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.8\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/KE_a08_loose}
    \caption{Evolution of the total kinetic energy.}
    \label{fig:KE_a08_loose}
\end{subfigure}
}\\
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.95\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/KExy_a08_loose}
    \caption{Evolution of horizontal and vertical kinetic energies.}
    \label{fig:KExy_a08_loose}
\end{subfigure}
}
\caption{Effect of permeability on the evolution of kinetic energies with time 
for a granular column collapse in fluid (a = 0.8 \& loose packing).}
\label{fig:a08_loose_energy}
\end{figure}

\begin{figure}
\centering
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.95\linewidth}
	\centering
    \includegraphics[width=0.8\linewidth]{figs/Packing_Density_a08_loose}
    \caption{Evolution of packing density.}
    \label{fig:Packing_Density_a08_loose}
\end{subfigure}
}\\
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.95\linewidth}
	\centering
    \includegraphics[width=0.8\linewidth]{figs/Fr_a08_loose}
    \caption{Evolution of Froude's number.}
    \label{fig:Fr_a08_loose}
\end{subfigure}
}
    \caption{Effect of permeability on the evolution of packing density and 
    Froude's number for a granular column collapse in fluid (a = 0.8 \& loose 
    initial packing).}
    \label{fig:Packing_Density_Fr_a08_loose}
\end{figure}

The evolution of grain trajectories with time are presented 
in~\cref{fig:Loose_a08_permeability} for low (\textit{r} = 0.95 \textit{R}) and 
high (\textit{r} = 0.9 \textit{R})
permeability conditions. It can be observed that a high 
permeability column shows a parabolic (convex) final profile in contrast to the 
more concave profile observed in low permeability condition, due to the effect 
of drag forces on the flow front. This difference in the flow thickness results 
in a higher value of Froude's number (0.59) and the occurrence of hydroplaning 
in the low permeability condition. Due to the high permeability, the water 
entrained at the flow front is dissipated quicker and 
thus no lubrication effect is observed. A Froude's number of 0.272 (no 
hydroplaning) is observed for the high permeability condition (\textit{r} = 0.7 
\textit{R}). The 
thick flow front in dense condition results in higher effective stress, in 
contrast to the low effective stress in loose condition due to positive 
pore-pressure at the flow front. The higher effective stress results in more 
frictional dissipation in dense condition, while the loose column experiences
lubrication effect. This shows that the drag force predominates at high 
permeability, while the low permeability condition is characterised by 
hydroplaning and lubrication.

\Cref{fig:effective_stress_a08} shows the normalised pressure at the 
base for the low and high permeability flows at $ t = 2\tau_c$. The normalised 
effective stress plotted is obtained as the average over 5 time steps at 
$2\tau_c$. The effective stress at the base is normalised to the effective 
stress of a static granular column before the collapse. A value of 1 indicates 
that the effective stress hasn't changed, which can be observed in the static 
region of the granular column. It can be observed that the normalised effective 
stress is significantly higher for the high permeability condition at the flow 
front in comparison to the almost non-existence of effective stress in the low 
permeability condition. The observation of trivial effective stress at the flow 
front corroborates the lubrication effect observed at low permeability 
conditions.

\begin{figure}
\centering
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/LBM_MD_loose_a08_r07_pt}
    \caption{High permeability (\textit{r} = 0.7 \textit{R}).}
    \label{fig:LBM_MD_loose_a08_r07_pt}
\end{subfigure}
}\\
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/LBM_MD_a08_Loose_r09}
    \caption{Low permeability (\textit{r} = 0.95 \textit{R}).}
    \label{fig:LBM_MD_a08_Loose_r09}
\end{subfigure}
}
\caption{Particle tracking of the deposit morphology
for a granular column collapse in fluid (a = 0.8 \& loose packing), influence 
of permeability.}
\label{fig:Loose_a08_permeability}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.97\columnwidth]{figs/a08/effective_stress_a08}
\caption{Effect of permeability on the normalised effective stress for loose 
initial packing at $t = 2\tau_c$.}
\label{fig:effective_stress_a08}
\end{figure}

\Cref{fig:Dense_Loose_PT} shows the grain trajectories of a dense and a loose 
initial packing for a hydrodynamic radius (\textit{r} = 0.95 \textit{R}). It 
can be observed that 
the dense initial packing results in a lot of turbulent behaviour at the flow 
surface in contrast to the more uniform flow behaviour in the loose condition. 
The thickness of the deposit in both dense and loose condition is almost the 
same, however the density of the flow results in a Froude's number of 0.59 and
0.429 for loose and dense conditions, respectively. The low initial density 
results in more hydroplaning in the loose condition. The effect of water 
entrainment at the flow front in dense and loose conditions can be seen 
in~\cref{fig:Dense_Loose_voro}. Water entrainment at the flow front can be 
observed in the loose condition, this is shown by white-coloured (empty Voronoi 
cells) at the flow front. This empty region in the granular packing between the 
granular mass and the base at the flow front represents the entrained water, 
which results in hydroplaning. Comparing the evolution packing densities in 
dense and loose conditions
(\cref{fig:Packing_Density_a08_dense,fig:Packing_Density_a08_loose}) 
reveal almost the same packing density when the flow is fully mobilised. Hence, 
it is the density of the flowing granular mass that controls the influence of 
hydroplaning for a given hydrodynamic radius and initial aspect ratio. A 
loosely packed granular column with low permeability entrains 
more water at the flow front, resulting in a hydroplaning effect that overcomes 
the influence of viscous drag forces and thereby yields a higher run-out 
distance than the dry condition.

\citeNP{Rondon2011} also observed that the collapse of a granular column in a 
viscous fluid is mainly controlled by the initial volume fraction and not by 
the aspect ratio of the column. The role of the initial volume fraction 
observed explains the pore pressure feedback mechanism proposed 
by~\citeNP{Schaeffer2008,Iverson2000} in the context of landslides. The 
compaction or dilation of grains can cause additional stress in the grains 
which can stabilise or destabilise the soil. The flow is thus controlled by the 
coupling between the dilatancy of the granular layer and the development of 
pore pressure in the fluid phase~\cite{Pailha2008}. The dense column needs to 
dilate in order to flow. When it starts to fall, liquid is then sucked into the 
column, which is then stabilised by the additional viscous 
drag~\cite{Rondon2011,Topin2012}. By opposition the loose column when 
it starts flowing expands and ejects liquid, leading to a partial fluidisation 
of the material.

\begin{figure}[htbp]
\centering
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/Dense_a08_voro_tc}
    \caption{Dense initial packing (83\%)}
    \label{fig:Dense_a08_voro_tc}
\end{subfigure}
}\\
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/Loose_a08_voro_tc}
    \caption{Loose initial packing (79\%)}
    \label{fig:Loose_a08_voro_tc}
\end{subfigure}
}
\caption{Evolution of packing fraction at $t = \tau_c$ for dense and loose 
initial packing fraction. Black means dense packing, while white colour denotes 
loose packing in the Voronoi cell.}
\label{fig:Dense_Loose_voro}
\end{figure}

\begin{figure}[htbp]
\centering
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.8\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/PT_Dense_R095}
    \caption{Dense initial packing (83\%)}
    \label{fig:PT_Dense_R095}
\end{subfigure}
}\\
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.8\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/PT_Loose_R095}
    \caption{Loose initial packing (79\%)}
    \label{fig:PT_Loose_R095}
\end{subfigure}
}
\caption[Effect of initial density on the deposit morphology
for a granular column collapse in fluid (a = 0.8).]{Effect of initial density 
on the deposit morphology
for a granular column collapse in fluid (a = 0.8). Dense vs. loose initial 
packing fraction (\textit{r} = 0.95 \textit{R}). Darker means dense packing, 
white indicates 
loose 
packing density.}
\label{fig:Dense_Loose_PT}
\end{figure}
